A Mountaintop View Requires Minimal Sorting: \break An Instance Optimal Contour Tree Algorithm

### 2014-10-27 paint-contour-tree.tex
intro changes

Here we present the first contour tree algorithm which takes into account the shape of the contour tree, 
and in doing so yields instance optimal results (on all instances requiring $\Omega(n\alpha(n))$ time).
In particular, our algorithm runs in $O(n\alpha(n))$ time when the tree is balanced, yet still 
matches the previous worst case $O(n\log n)$ bound when the tree is long and skinny.

@@

There is a long list of practical applications of the study of contour trees (see for example \cite{c-tmi-04} and references therein).
Often these applications involve massive data sets, and so consequently there has also been a long series 
of results on efficient algorithms for their computation \cite{kobps-ctsssit-97,tv-cct-98,csa-cctad-00,pc-ectls-02,cllr-sooscctmp-05}.
As an initial step these algorithms require that all $n$ contour tree vertices be sorted.  
In particular, \cite{cllr-sooscctmp-05} provide an $O(n\log n)$ time algorithm, and since in the worst case computing the contour tree 
requires $\Omega(n\log n)$ time, this algorithm is optimal with respect to the size of the output.  
Here we present the first contour tree algorithm which is optimal (for instances requiring $\Omega(n\alpha(n))$ 
time\footnote{Where $\alpha(n)$ is the inverse Ackermann function, which is $< 5$ for computation in this universe.}) 
with respect not just to the size of the output, but also its structure.  This is accomplished by introducing a simple linear time 
surface partitioning scheme to avoid sorting, which is likely to have applications elsewhere.  
For all instances the bound for our algorithm either matches the best previous bound, or is faster.
For, example when the contour tree is a balanced, our algorithm runs in $O(n\alpha(n))$ time, whereas all 
previous algorithms had a running time of $\Omega(n\log n)$

@@

There have also been some output sensitive results for computing the contour tree.  Let $t'$ be the number 
of critical points in the input mesh, which could potentially be much smaller than $n$.  
Cole-McLaughlin and Pascucci \cite{pc-ectls-02} provided an $O(n+t'\log n)$ time algorithm for the case of 
$3$-dimensional structured meshes (note $N$ does not show up in the running time as the mesh is structured).
Chiang \etal \cite{cllr-sooscctmp-05} improve upon this result, and provide an $O(N+t\log t)$ algorithm 
which works for structured or unstructured meshes in any dimension, where $t$ is the number of component 
critical vertices.  This is provably optimal with respect to the output size but not the output structure, 
as discussed below.

@@

 The benefit of this partitioning is that each resulting piece 
 is maxima dominant, i.e. there is a single maxima that has a non-ascending path to each point.  
 This implies the contour of this submanifold is the same as its split tree, or in other words 
 its contour tree is a rooted binary tree.

@@

\paragraph{Lower Bound Overview.}
\Ben{To be Written}

@@

We use $\cV$ to denote the set of critical points in $\MM$.
Because $f$ is piecewise-linear, all critical points are vertices in $\MM$. 
A value $h$ is called critical, if $f(x) = h$, for some $x \in \cV$. The Morse assumption
implies that all critical points have distinct values. A contour that involves a critical
point is also denoted as critical. Note that the vertices in a boundary face collectively behaves like an extremum.
Abusing notation, the term ``critical point", ``maxima", and ``minima" will include boundaries.
Critical values include these boundary values. By boundary criticality, all contours are themselves
represented as $d-1$-dimensional simplicial complices.

@@

\begin{definition} \label{join} The join tree $\cJ(\MM)$ of $\MM$ is built on vertex set $V(\MM)$.
The directed edge $(u,v)$ is present when $u$ is the smallest valued vertex is a connected component of $\MM^+_v$
\emph{and} $v$ is connected to this component (in $\MM$). The split tree $\cS(\MM)$ is obtained
by looking at $\MM^-_v$ (or alternately, by taking the join tree of the inversion of $\MM$). 
\end{definition}

@@

\begin{proof} First, let us complete the proof assuming that each split $v$ has one child that is just a leaf, which
is a non-dominant minimum. Any minimum is a leaf in $\cS(\MM)$, and thereby attached to a split. If we removed
all non-dominant minima, we must end up with a path, as asserted above.

Consider a split $v$. For sufficiently small $\eps, \delta$, there are exactly two $(f(v) - \delta)$-contours
$\phi$ and $\psi$ intersecting $B_\eps(v)$. Both of these are regular contours. There must be a non-ascending
path from $v$ to the dominant minimum $m^*$. Consider the first edge (necessarily decreasing from $v$)
on this path. It must intersect some contour, say $\phi$. By \Thm{jordan}, $\cut(\MM,\phi)$ has
two connected components, with one (call it $\LL$) having $\phi^-$ as a boundary maximum. This complex
contains $m^*$ as the non-ascending path $\phi$ intersects $\phi$ only once. Let the other
component be called $\MM'$.

@@

Now we come to the main algorithmic contribution. We design a new algorithm for computing contour
trees of a minimum dominant simplicial complice $\MM$. As part of the initialization, we determine
all critical points in $\MM$. For any vertex $v$, we can look at $f(v) \pm \delta$-contours 
in $f|_{B_\eps(v)}$ to determine the up and down degrees. This determines the criticality (and type)
of $v$. This requires $O(|\MM|)$ time, since every facet is looked at proportional to the number of its
vertices.

{\bf Painting:} The central tool is a notion 
of \emph{painting} of $\MM$. Initially associate a color with each maximum. Denoting the set of maxima is $X$,
this is an arbitrary bijection $\chi:X \mapsto [|X|]$. Imagine there being a large
can of paint of color $\chi(x)$ at maxima $x$. We will spill different paint from each maximum and watch it flow down.
This is analogous to the raining in the previous section, but paint is a much more viscous liquid.
\emph{So paint only flows down edges, and it does not color the interior of facets.} Furthermore, paints
do not mix, so eventually, every vertex and edge of $\MM$ gets a unique color. This process (and indeed the entire algorithm)
works purely on the 1-skeleton of $\MM$, which is just a graph.
%
\begin{definition} \label{def:paint} Let the 1-skeleton of $\MM$ be denoted $G(\MM) = (\cV,E)$.
A  \emph{painting} of $\MM$ is a map $\chi:X \cup E \mapsto [|X|]$ with the following property. 
 Consider an edge $e$. There exists a descending path from some maximum $x$ to $e$
	consisting of edges in $E$, such that all edges along this path have the same color as $x$. (The maximum $x$
	is well-defined to be the unique maximum with the same color as $e$.)

An \emph{initial} painting has the additional restriction that the restriction $\chi:X \mapsto [|X|]$ is a bijection.
\end{definition}
%
Paintings are not unique. It is straightforward to construct an initial painting of $\MM$ in $O(|\MM|)$ time, 
using a descending BFS from each maximum that does not explore previously colored edges. 
As the algorithm proceeds, the bijection property will not hold, though we always maintain a valid painting.

@@

\begin{definition} \label{def:color-full} Consider vertex $v$. An \emph{up-star} of $v$ is the set 
of edges that all connected to a fixed component of $\MM^+_v$. (Each $v$
participates in at mot $2$ up-stars. A color $c \in \col(v)$ \emph{fully touches} $v$
if all edges in an up-star are colored $v$.
%and let $\phi$ be a $(f(v)+\delta)$-contour in
%$f|_{B_\eps(v)}$. If $\phi$ is monochromatic of color $c$, then $v$ is \emph{fully touched by $c$}. The set of colors
%that fully touch $v$ is denoted $\fcol(v)$.
\end{definition}

%Note that $\fcol(v)$ has size at most $2$, by the Morse condition. 
Note that $\col(v)$ 
%and $\fcol(v)$ can
be initially determined in $O(|\MM|)$ time, by a simple linear scan.

@@

\begin{proof} Since $e$ has color $c$,
there must exist vertices in $\MM^+_v$ touched by $c$. Consider the highest
vertex $w$ in $\MM^+_v$ that is touched by $c$ and some other color. If no such vertex exists,
this means all edges incident the a vertex touched by $c$ are colored $c$. By walking through
$\MM^+_v$, we deduce that all edges are colored $c$.

So assume $w$ exists. Take the $(f(w)+\delta)$-contour $\phi$ that intersects $B_\eps(v)$
and intersect some $c$-colored edge incident to $w$. Note that all edges intersecting $\phi$ are also colored $c$,
since $w$ is the highest vertex to be touched by $c$ and some other color. (Take the path of $c$-colored
edges from the maximum to $w$. For any point on this path, the contour passing through this point must
be colored $c$.) Hence, $c$ fully touches $w$. 
But $w$ is touched by another color, and the corresponding edge cannot intersect $\phi$. So $w$
must have up-degree $2$ and is critical.
\end{proof}

@@

During this process, various colors will be \emph{merged}. This involves merging the respective $T(c)$'s and
updating the union-find data structure. The only operation performed on $T(c)$'s are delete-max and merge.
No new vertex is ever inserted into the heaps. As colors merge, the initial painting is obviously modified.
But it would be too expensive to explicitly update colors. So the new painting is given by $\rep(c)$,
while we only have access to the original painting. By union-find bounds, the only adds an $\alpha(|\MM|)$
overhead to the data structure operations. The new coloring is guaranteed to be a painting of $\NN$.

While this is the intuitive viewpoint, we describe how this information is actually maintained.
The vertices $\{\h(\rep(c))\}$ represent the set of contours $\Phi$. In terms of the contour tree, 
we have already constructed the contour trees rooted at each $h(\rep(c))$. This is trivially true at the initialization,
when each $h(c)$ is a maximum. One can now treat each $h(\rep(c))$ as a maximum, corresponding to (a possibly new) color $\rep(c)$.
To reduce confusion between the original painting and the new colors, we will assume that $\col(v)$ and $\fcol(v)$
are as given by the original painting.

>>>>

Here we present the first contour tree algorithm which takes into account the shape of the contour tree, 
and in doing so yields instance optimal results.
Specifically, we prove that for any partition of the contour tree into a set of descending paths, $P$, 
any comparison based algorithm requires $\sum_{p\in P} |p|\log|p|$ time.  Moreover, we show there   
exists some partition $P'$ such that the running time of our algorithm is 
$O(\sum_{p\in P'} |p|\log|p| + N\alpha(N))$, where $N$ is total complexity of the input manifold 
(and hence this is optimal on all instances requiring $\Omega(N\alpha(N))$ time).
In particular, our algorithm runs in $O(N\alpha(N))$ time when the tree is balanced, 
while previous algorithms ran in $O(N+t\log t)$ (where $t$ is the size of the contour tree).
%In particular, our algorithm runs in $O(N\alpha(N))$ time when the tree is balanced, yet still 
%matches the previous worst case $O(N+n\log n)$ bound when the tree is long and skinny.

@@

There is a long list of practical applications of the study of contour trees (see for example \cite{c-tmi-04} and references therein).
Often these applications involve massive data sets, and so consequently there has also been a long series 
of results on efficient algorithms for their computation \cite{kobps-ctsssit-97,tv-cct-98,csa-cctad-00,pc-ectls-02,cllr-sooscctmp-05}.
Let $N$ be the total size of the input manifold, $n$ the number of vertices, and $t$ the number of critical points. (Note that in many applications $N=O(n)$.)
As an initial step these algorithms require that all $t$ contour tree vertices be sorted.  
In particular, \cite{cllr-sooscctmp-05} provide an $O(N+t\log t)$ time algorithm, and since in the worst case computing the contour tree 
requires $\Omega(N+t\log t)$ time, this algorithm is optimal with respect to the size of the output.  
Here we present the first contour tree algorithm which is optimal (for instances requiring $\Omega(N\alpha(N))$ 
time\footnote{Where $\alpha(\cdot)$ is the inverse Ackermann function, which is $< 5$ for computation in this universe.}) 
with respect not just to the size of the output, but also its structure.  This is accomplished by introducing a simple linear time 
surface partitioning scheme to avoid sorting, which is likely to have applications elsewhere.  
For all instances the bound for our algorithm either matches the best previous bound, or is faster.
For, example when the contour tree is a balanced, our algorithm runs in $O(N\alpha(N))$ time, whereas all 
previous algorithms had a running time of $\Omega(N+t\log t)$

@@

There have also been some output sensitive results for computing the contour tree.  Let $t$ be the number 
of critical points in the input mesh, which could potentially be much smaller than $n$.  
Cole-McLaughlin and Pascucci \cite{pc-ectls-02} provided an $O(n+t\log n)$ time algorithm for the case of 
$3$-dimensional structured meshes (note $N$ does not show up in the running time as the mesh is structured).
Chiang \etal \cite{cllr-sooscctmp-05} improve upon this result, and provide an $O(N+t'\log t')$ algorithm 
which works for structured or unstructured meshes in any dimension, where $t'$ is the number of component 
critical vertices.  This is provably optimal with respect to the output size but not the output structure, 
as discussed below.

@@

 The benefit of this partitioning is that each resulting piece 
 is maxima dominant, i.e. there is a single maxima that has a non-ascending path to each point.  
 This implies the contour tree of this submanifold is the same as its split tree, or in other words 
 its contour tree is a rooted binary tree.

@@

\paragraph{Lower Bound Overview.}
In order to show that our algorithm for computing the join tree is optimal we consider paritioing the tree into disjoint 
descending paths, each containing a leaf.  It turns out that each such partition into paths, $P$, 
defines a lower bound of $\sum_{p\in P} |p|\log|p|$ for computing the contour tree, as you need to know the 
sorted order of the vertices along each path in order to determine the tree.
Specifically, we show that there is a set of at least $\Pi_{p_i\in P} (|p_i|-1)!$ distinct PL $2$-manifolds, 
all of which have the same defining mesh, and for which each sorted order of each group of saddles determines a different join tree.

This instance optimality of our algorithm then follows, as we also show that there is a particular path decomposition 
whose cost upper bounds our.  Specifically, we assign a weight to each node in the tree which is determined by 
the painting procedure described above (which represents the time it takes to process this node).  
The path decomposition then turns out to be the heavy-light path decomposition 
described by Sleator and Tarjan \cite{st-dsdt-83}, according to this weight function.
However, this decomposition is just the first step in the analysis, which 
requires further case analysis depending on the relative size of the length of each path 
and the weights of its vertices.

@@

We use $\cV$ to denote the set of critical points in $\MM$.
Because $f$ is piecewise-linear, all critical points are vertices in $\MM$. 
A value $h$ is called critical, if $f(x) = h$, for some $x \in \cV$. The Morse assumption
implies that all critical points have distinct values. A contour that involves a critical
point is also denoted as critical. Note that the vertices in a boundary face collectively behaves like an extremum.
Abusing notation, the term ``critical point", ``maxima", and ``minima" will include boundaries.
Critical values include these boundary values. By boundary criticality, all contours are themselves
represented as $d-1$-dimensional simplicial complexes.

@@

\begin{definition} \label{join} The join tree $\cJ(\MM)$ of $\MM$ is built on vertex set $V(\MM)$.
The directed edge $(u,v)$ is present when $u$ is the smallest valued vertex in a connected component of $\MM^+_v$
\emph{and} $v$ is connected to this component (in $\MM$). The split tree $\cS(\MM)$ is obtained
by looking at $\MM^-_v$ (or alternately, by taking the join tree of the inversion of $\MM$). 
\end{definition}

@@

\begin{proof} It suffices to prove that each split $v$ has one child that is just a leaf, which
is a non-dominant minimum.  Specifically, any minimum is a leaf in $\cS(\MM)$ and thereby attached to a split, 
which implies that if we removed all non-dominant minima, we must end up with a path, as asserted above.

Consider a split $v$. For sufficiently small $\eps, \delta$, there are exactly two $(f(v) - \delta)$-contours
$\phi$ and $\psi$ intersecting $B_\eps(v)$. Both of these are regular contours. There must be a non-ascending
path from $v$ to the dominant minimum $m^*$. Consider the first edge (necessarily decreasing from $v$)
on this path. It must intersect one of the $(f(v) - \delta)$-contours, say $\phi$. By \Thm{jordan}, $\cut(\MM,\phi)$ has
two connected components, with one (call it $\LL$) having $\phi^-$ as a boundary maximum. This complex
contains $m^*$ as the non-ascending path intersects $\phi$ only once. Let the other
component be called $\MM'$.

@@

Now we come to the main algorithmic contribution. We design a new algorithm for computing contour
trees of a minimum dominant simplicial complex $\MM$. As part of the initialization, we determine
all critical points in $\MM$. For any vertex $v$, we can look at $f(v) \pm \delta$-contours 
in $f|_{B_\eps(v)}$ to determine the up and down degrees. This determines the criticality (and type)
of $v$. This requires $O(|\MM|)$ time, since every facet is looked at proportional to the number of its
vertices.

{\bf Painting:} The central tool is a notion 
of \emph{painting} of $\MM$. Initially associate a color with each maximum. Denoting the set of maxima by $X$,
this is an arbitrary bijection $\chi:X \mapsto [|X|]$. Imagine there being a large
can of paint of color $\chi(x)$ at maxima $x$. We will spill different paint from each maximum and watch it flow down.
This is analogous to the raining in the previous section, but paint is a much more viscous liquid.
\emph{So paint only flows down edges, and it does not color the interior of facets.} Furthermore, paints
do not mix, so eventually, every vertex and edge of $\MM$ gets a unique color. This process (and indeed the entire algorithm)
works purely on the 1-skeleton of $\MM$, which is just a graph.
%
\begin{definition} \label{def:paint} Let the 1-skeleton of $\MM$ be denoted $G(\MM) = (\cV,E)$.
A  \emph{painting} of $\MM$ is a map $\chi:X \cup E \mapsto [|X|]$ with the following property. 
 Consider an edge $e$. There exists a descending path from some maximum $x$ to $e$
	consisting of edges in $E$, such that all edges along this path have the same color as $x$. (The maximum $x$
	is well-defined to be the unique maximum with the same color as $e$.)

An \emph{initial} painting has the additional property that the restriction $\chi:X \mapsto [|X|]$ is a bijection.
\end{definition}
%
Paintings are not unique. It is straightforward to construct an initial painting of $\MM$ in $O(|\MM|)$ time, 
using a descending BFS from each maximum that does not explore previously colored edges. 
As the algorithm proceeds, the bijection property will not hold, though we always maintain a valid painting.

@@

\begin{definition} \label{def:color-full} Consider vertex $v$. An \emph{up-star} of $v$ is the set 
of edges that all connect to a fixed component of $\MM^+_v$. (Each $v$
participates in at mot $2$ up-stars.) A color $c \in \col(v)$ \emph{fully touches} $v$
if all edges in an up-star are colored $v$.
%and let $\phi$ be a $(f(v)+\delta)$-contour in
%$f|_{B_\eps(v)}$. If $\phi$ is monochromatic of color $c$, then $v$ is \emph{fully touched by $c$}. The set of colors
%that fully touch $v$ is denoted $\fcol(v)$.
\end{definition}

%Note that $\fcol(v)$ has size at most $2$, by the Morse condition. 
Note that $\col(v)$ 
%and $\fcol(v)$ can
can be initially determined in $O(|\MM|)$ time, by a simple linear scan.

@@

\begin{proof} Since $e$ has color $c$,
there must exist vertices in $\MM^+_v$ touched by $c$. Consider the highest
vertex $w$ in $\MM^+_v$ that is touched by $c$ and some other color. If no such vertex exists,
this means all edges incident to a vertex touched by $c$ are colored $c$. By walking through
$\MM^+_v$, we deduce that all edges are colored $c$.

So assume $w$ exists. Take the $(f(w)+\delta)$-contour $\phi$ that intersects $B_\eps(v)$
and intersect some $c$-colored edge incident to $w$. Note that all edges intersecting $\phi$ are also colored $c$,
since $w$ is the highest vertex to be touched by $c$ and some other color. (Take the path of $c$-colored
edges from the maximum to $w$. For any point on this path, the contour passing through this point must
be colored $c$.) Hence $c$ fully touches $w$. 
But $w$ is touched by another color, and the corresponding edge cannot intersect $\phi$. So $w$
must have up-degree $2$ and is critical.
\end{proof}

@@

During this process, various colors will be \emph{merged}. This involves merging the respective $T(c)$'s and
updating the union-find data structure. The only operation performed on $T(c)$'s are delete-max and merge.
No new vertex is ever inserted into the heaps. As colors merge, the initial painting is obviously modified.
But it would be too expensive to explicitly update colors. So the new painting is given by $\rep(c)$,
while we only have access to the original painting. By union-find bounds, this only adds an $\alpha(|\MM|)$
overhead to the data structure operations. The new coloring is guaranteed to be a painting of $\NN$.

%While this is the intuitive viewpoint, we describe how this information is actually maintained.
The vertices $\{\h(\rep(c))\}$ represent the set of contours $\Phi$. In terms of the contour tree, 
we have already constructed the contour trees rooted at each $h(\rep(c))$. This is trivially true at the initialization,
when each $h(c)$ is a maximum. One can now treat each $h(\rep(c))$ as a maximum, corresponding to (a possibly new) color $\rep(c)$.
To reduce confusion between the original painting and the new colors, we will assume that $\col(v)$ and $\fcol(v)$
are as given by the original painting.
